Denotational Semantics of General Payment Primitives, and Its Payment System

### 2022-09-30 packages/spec-haskell/yellowpaper/Paper.tex
complete (mostly) Denotational Semantics section

To come up with such lawful money distribution, we can divide and conquer. In a payment system, if we restrict that only
payment primitives can provide ``updates'' to its money distribution, so that a money distribution is only a result of a
sequence of payment primitive updates; then we can have these axioms as the basis for the proof:

\centerline{\textbf{Axioms of restricted money distributions}}

\begin{equation}
    \begin{split}
        &\textbf{Axiom A:\ }
        \textit{law of conservation of value for payment primitives} \\
        &\forall t \in \mathbb{R} ({\displaystyle \sum_{u \in U} [\![prim\ args]\!]\ u\ t = 0}) \\
        \\
        &\textbf{Axiom B:\ }
        \textit{money distribution consists of updates from payment primitives} \\
        &[\![m]\!] = [\![prim_1\ args_1]\!] \oplus [\![prim_2\ args_2]\!] \oplus [\![prim_3\ args_3]\!] \oplus \dotsb
    \end{split}
\end{equation}

This allows us to impose the law of conservation of value on payment primitives, then prove inductively that such
restricted money distribution obey the law of conservation of value too.

\centerline{\textbf{Proof of restricted money distributions satisfying the law of conservation of value}}

\begin{equation}
    \begin{split}
        &\textbf{In addition to the axioms, given:} \\
        \\
        &\textit{(base case with empty money distribution is trivial)} \\
        &\forall t \in \mathbb{R} {\displaystyle \sum_{u \in U} [\![\emptyset]\!]\ u\ t = 0} \\
        \\
        &\textit{(monoid homomorphism)} \\
        &{\displaystyle \sum_{u \in U} [\![prim_1\ args_1\oplus prim_2\ args_2]\!]\ u\ t} = \\
        &\qquad {\displaystyle \sum_{u \in U} [\![prim_2\ args_2]\!]\ u\ t} \oplus
          {\displaystyle \sum_{u \in U} [\![prim_2\ args_2]\!]\ u\ t} \\
        \\
        &\textbf{We have:} \\
        &\forall t \in \mathbb{R}. \\
        &{\displaystyle \sum_{u \in U} [\![m]\!]\ u\ t} \\
        &\textit{(applying Axioms B)} \\
        = &{\displaystyle \sum_{u \in U}
            ([\![prim_1\ args_1]\!] \oplus
            [\![prim_2\ args_2]\!] \oplus
            [\![prim_3\ args_3]\!]\oplus \dotsb)
        }\ u\ t\\
        &\textit{(applying monoid homomorphism)} \\
        = &{\displaystyle \sum_{u \in U} [\![prim_1\ args_1]\!]\ u\ t} \oplus
        {\displaystyle \sum_{u \in U} [\![prim_2\ args_2]\!]\ u\ t} \oplus
        {\displaystyle \sum_{u \in U} [\![prim_3\ args_3]\!]\ u\ t} \oplus \dotsb
        )\\
        &\textit{(applying Axioms A)} \\
        =&\ 0
        \\
        \blacksquare
    \end{split}
\end{equation}

@@

@@

A \textit{index} has a function $\rho$ which produces a real number for each monetary unit:

\begin{equation}
    \rho :: k \rightarrow u \rightarrow \mathbb{R}
\end{equation}

@@

\begin{equation}
    \displaystyle \sum_{u \in U} \rho\ u = 1
\end{equation}

@@

Now let's generalize the one to one payment primitives using index abstraction. We use symbol $k$ for the indexes, and
$I$ subscript for the indexed versions of the primitives.

@@

@@

Now one to one payment primitives can be redefined using \textit{universal index} ($ku_{any}$) define as such:

@@

It means that it's an index universally available for any monetary unit, and its proportion is always 1 for that
monetary unit.

@@

\begin{equation}
    unit :: sub \rightarrow subs \rightarrow Int
\end{equation}

\begin{equation}
    proportion\ sub\ subs = {{unit\ sub\ subs } \over {\displaystyle \sum_{x \in subs} unit\ x\ subs}}
\end{equation}

\begin{equation}\label{sem_distributeProportionally}
    \begin{split}
        [\![&distributeProportionally\ pub\ subs\ amount\ md]\!] = [\![md]\!]\ \oplus \\
        (\lambda\ case&\ mu\ t \\
        &|\ pub = mu \rightarrow -amount \\
        &|\ \exists sub \in subs\ {sub = mu } \rightarrow amount * (proportion\ mu\ subs) \\
        &|\ otherwise \rightarrow 0 \\
        )
    \end{split}
\end{equation}

\begin{equation}\label{sem_distributeProportionally}
    \begin{split}
        [\![&updateProportionalDistributionConstantFlow\ pub\ subs\ flowRate\ t'\ md]\!] = [\![md]\!]\ \oplus \\
        (\lambda\ case&\ mu\ t \\
        &|\ pub = mu \rightarrow -flowRate * (t - t') \\
        &|\ \exists sub \in subs\ {sub = mu } \rightarrow flowRate * (t - t') * (proportion\ mu\ subs) \\
        &|\ otherwise \rightarrow 0 \\
        )
    \end{split}
\end{equation}

>>>>

To come up with such lawful money distribution, we can divide and conquer. In a payment system, if we restrict that only
payment primitives can provide ``updates'' to its money distribution, so that a money distribution is only a result of a
sequence of payment primitive updates; then we can have these axioms as the basis for the proof.

\paragraph{Axiom A of Restricted Money Distributions}

First we must impose the law of conservation of value on payment primitives as an axiom, in order to prove inductively
that such restricted money distribution satisfies the law of conservation of value too.

\begin{equation}
    \begin{split}
        &\textit{law of conservation of value for payment primitives} \\
        &\forall t \in \mathbb{R} ({\displaystyle \sum_{u \in U} [\![prim\ args]\!]\ u\ t = 0}) \\
    \end{split}
\end{equation}

\paragraph{Axiom B of Restricted Money Distributions}

\begin{equation}
    \begin{split}
        &\textit{money distribution consists of updates from payment primitives} \\
        &[\![m]\!] = [\![prim_1\ args_1]\!] \oplus [\![prim_2\ args_2]\!] \oplus [\![prim_3\ args_3]\!] \oplus \dotsb
    \end{split}
\end{equation}

\paragraph{Proof of Restricted Money Distributions}

satisfying the law of conservation of value.

\begin{equation}
    \begin{split}
        &\textbf{In addition to the axioms, given:} \\
        \\
        &\textit{(base case with empty money distribution is trivial)} \\
        &\forall t \in \mathbb{R} {\displaystyle \sum_{u \in U} [\![\emptyset]\!]\ u\ t = 0} \\
        \\
        &\textit{(monoid homomorphism)} \\
        &{\displaystyle \sum_{u \in U} [\![prim_1\ args_1\oplus prim_2\ args_2]\!]\ u\ t} = \\
        &\qquad {\displaystyle \sum_{u \in U} [\![prim_2\ args_2]\!]\ u\ t} \oplus
          {\displaystyle \sum_{u \in U} [\![prim_2\ args_2]\!]\ u\ t} \\
        \\
        &\textbf{We have:} \\
        &\forall t \in \mathbb{R}. \\
        &{\displaystyle \sum_{u \in U} [\![m]\!]\ u\ t} \\
        &\textit{(applying Axiom B)} \\
        = &{\displaystyle \sum_{u \in U}
            ([\![prim_1\ args_1]\!] \oplus
            [\![prim_2\ args_2]\!] \oplus
            [\![prim_3\ args_3]\!]\oplus \dotsb)
        }\ u\ t\\
        &\textit{(applying monoid homomorphism)} \\
        = &{\displaystyle \sum_{u \in U} [\![prim_1\ args_1]\!]\ u\ t} \oplus
        {\displaystyle \sum_{u \in U} [\![prim_2\ args_2]\!]\ u\ t} \oplus
        {\displaystyle \sum_{u \in U} [\![prim_3\ args_3]\!]\ u\ t} \oplus \dotsb
        )\\
        &\textit{(applying Axiom A)} \\
        =&\ 0
        \\
        \blacksquare
    \end{split}
\end{equation}

@@

It should be easy to prove that they satisfy the \textit{axioms A of restricted money distribution}.

@@

A \textit{index} (we use symbol $k$ for them) has a function $\rho$ which produces a real number for each monetary unit:

\begin{equation}
    \rho\ k :: u \rightarrow \mathbb{R}
\end{equation}

@@

\begin{equation}
    \displaystyle \sum_{u \in U} \rho\ k\ u = 1
\end{equation}

@@

\subsection{Indexed Primitives}

Let's generalize the one to one payment primitives using index abstraction. We use $I$ subscript for the indexed
versions of the primitives.

@@

It should be straightforward to prove that they also satisfy the \textit{axiom A of restricted money distribution}
thanks to the law of the index.

@@

Now one to one payment primitives can be redefined using \textit{universal index} ($ku_{any}$):

@@

It means that it is an index \textit{universally} available for each monetary unit, and its proportion is always 1 for
that monetary unit and 0 for all others.

@@

A special case of the indexed primitives is to fix the sender side to be an \textit{universal index}.

The semantics of these primitives resembles dividend pay out (distribution) in the security market\footnote{What Is a
Dividend?  https://www.investopedia.com/terms/d/dividend.asp}. And flow semantics enables a whole new type of dividend
pay out that is continuous over time.

\subsection{Network Abstraction}

An even more powerful abstraction is to model participants involved a payment primitive a \textit{network} (we use
symbol $w$ for them).

It has the function $\rho$ as in \textit{index}, bnd it satisfies a different law:

\begin{equation}
    \displaystyle \sum_{u \in U} \rho\ w\ u = 0
\end{equation}

\subsection{Networked Primitives}

Let's generalize the basic payment primitives using network abstraction. We use $N$ subscript for the indexed versions
of the primitives.

\paragraph{Shift}

We rename $transfer$ to $shift$ for networked instant payment primitive:

\begin{equation}
    \begin{split}
        [\![shift_N\ w\ x]\!] &=
        \lambda u \rightarrow \lambda t \rightarrow x \cdot \rho\ w\ u
    \end{split}
\end{equation}

\paragraph{Networked (Constant) Flow}

\begin{equation}
    \begin{split}
        [\![flow_N\ w\ r\ t']\!] &=
        \lambda u \rightarrow \lambda t \rightarrow r \cdot (t - t') \cdot \rho\ w\ u
    \end{split}
\end{equation}

\subsection{Generality vs. Optimization}

The question now is, should we use the most general form of the semantics to guide implementation?

The answer is most likely a no. Not because it is not useful, after all mathematical formula is about truth and perhaps
also elegance in it, but because of we may miss optimization opportunities needed in implementation when more
specialized versions are used instead\footnote{We will not though discuss the subjective aspects, including in software
engineering principles such as YAGNI (\url{https://en.wikipedia.org/wiki/You_aren't_gonna_need_it}), or user experience
perspective of the payment primitives.}.

In the next chapter, we will look into a reference implementation where these optimizations are applied.
